\bfseries\large Phosphatidylinositol 4,5-bisphosphate regulates cilium transition zone maturation in \textit{Drosophila melanogaster}

### 2018-02-20 latex/paper.tex
update
updated

\section{Introduction}
Cilia are sensory organelles present on almost all cells in the human body
\citep{satir2010primary}.
They function primarily by providing sensory inputs to cells, but can
also function as motile propellers in certain cells
\citep{bloodgood2010sensory}.
Consistent with their pervasiveness and importance in cell signalling,
defects in cilium formation are associated with various genetic disorders
known collectively as ciliopathies, which can 
neurological defects \citep{valente2014primary},
skeletal abnormalities \citep{hammarsjo2017novel, waters2011ciliopathies},
and infertility \citep{inaba2016sperm} in addition to other phenotypes.

A large number of ciliopathies are linked to mutations in proteins that
localize to the cilium transition zone (referred to herein as TZ).
The TZ is the proximal-most region of the cilium and
acts as a diffusion barrier that regulates the
bi-directional transport of protein and lipid cargo at the base of the cilium.
The TZ is essential for cilium formation (ciliogenesis) and cilium-associated
signalling.
The conserved TZ protein CEP290, for example, is mutated in at least
five different ciliopathies, and is essential for cilium formation in both humans
and Drosophila \citep{basiri2014migrating}.
Although the protein composition of TZs has been investigated in various
studies ranging from single gene analysis to proteomics,
processes underlying TZ maturation are not as well-studied.

In the Drosophila male germline, nascent TZs first assemble on basal bodies
during early G2 phase in spermatocytes and anchor directly to the plasma membrane.
This is followed by establishment of cilium-membrane connections
through a poorly understood mechanism
involving microtubule rearrangements \citep{gottardo2013cilium} and a
ciliary membrane that persists through meiosis \citep{riparbelli2012assembly}.
Ciliogenesis begins with the assembly of TZ proteins at the tips of
the nascent basal body.
This is followed by the anchoring of the TZ to the membrane and
establishment of a compartmentalized space bounded by the ciliary membrane
and the TZ.
TZ maturation has been described in
\textit{Paramecium} \citep{aubusson2015transition}
\textit{Caenorrhabditis elegans} \citep{serwas2017centrioles} and
\textit{Drosophila melanogaster} \citep{gottardo2013cilium}
and is most readily observed by an increase in TZ length from nascent to mature.
Possibly, this is similar to the initial steps of TZ assembly and reorganization
in well-studied vertebrate cilia.

Here, we show that the lipid phosphatidylinositol 4,5-bisphosphate (\PIP{})
is essential for proper TZ maturation in developing male germ cells in \textit{D. melanogaster}.
Reduction of cellular \PIP{} by expression of the bacterial PIP phosphatase SigD
or by inactivation of the PIP 5-kinase Skittles (Sktl) that localizes to cilia
induces longer than normal TZs without affecting TZ assembly.
These hyperelongated TZs show hallmarks of aberrant function similar to those
found in previously studied TZ mutants, including loss of cilium-membrane associations.
We also find that the \textit{onion rings} allele of the
ciliopathy-associated \textit{EXOC8} gene encoding the human Exo84
decouples TZ hyperelongation from loss of cilium-membrane contacts.
Our results suggest that the plasma membrane lipid \PIP{} regulates TZ length,
maturation and function, acting to support ciliogenesis.

@@

\subsection{The \textit{onion rings} mutant decouples defects found in cells with reduced levels of \PIP{}}
A heterozygous mutation in the human \textit{EXOC8},
the human ortholog of yeast Exo84, is linked to
Joubert syndrome \citep{dixon2012exome}.
EXOC8, the human ortholog of Exo84, is a component of the octameric exocyst complex, which binds
\PIP{} and regulates membrane trafficking at the plasma membrane.
We previously showed that male flies homozygous for the \textit{onion rings}
(\textit{onr}) mutant of Drosophila Exo84 were sterile, with phenotypes
similar to \sigd{} \citep{wei2008depletion}.
To investigate whether defects in TZ hyperelongation could be explained by
defective Exo84 function, we examined TZs in \textit{onr} mutants.
We found that the \textit{onr} mutant did not display hyperelongated TZs (Figure 5),
suggesting that Exo84 is dispensable for TZ maturation.

@@

\section{Discussion}
Cilia are essential for proper human development, and their disruption
can cause genetic disorders which display severity ranging from congenital
lethality to adult onset.
A large number of ciliopathies are associated with genes that encode for
proteins that function at the cilium transition zone (TZ).
Often, a single gene is linked to multiple ciliopathies with distinct phenotypes
and severities.
For example, \textit{CEP290} is linked to at least six different
cilium-associated disorders.
It is likely that these mutations do not completely abolish CEP290
function, and may be affected by genetic modifiers in an affected individual
\citep{zaghloul2010functional}.
Animal models with partial loss of TZ function may therefore provide
insight on intermediate phenotypes associated with human disease.
Here, we found that reduced levels of the membrane lipid \PIP{},
which is known to be enriched at the base of cilia,
affects TZ maturation, but not assembly,
leading to hyperelongated TZs that show signs of functional defects, and
may represent one such modifier for TZ function.

@@

%%
%    FIGURE 2
%
\textbf{Figure 2. Sktl is important for transition zone maturation and localizes to basal bodies.} 
\begin{enumerate}[label={(\Alph*)}, nolistsep]
\item Expression of full-length Sktl suppresses \sigd{}-induced TZ hyperelongation.
\item Quantification of Cep290 and Ana1 lengths from control, \sigd{} and Sktl; \sigd{} ($n$\textgreater 60) from (A).
\item \textit{sktl}$^{2.3}$ clones exhibit TZ hyperelongation.
\item Phylogenetic tree of PIP5Ks showing conservation of cilium-associated functions.
  The scale bar represents expected amino acid substitutions per site and branch support values are shown in red (a value of 1 indicates maximum support).
Abbreviations: Cele (\textit{Caenorhabditis elegans}), Spur (\textit{Strongylocentrotus purpuratus}), Amel (\textit{Apis mellifera}), Aaeg (\textit{Aedes aegypti}), Dana (\textit{Drosophila ananassae}), Dmel (\textit{Drosophila melanogaster}), Hsap (\textit{Homo sapiens}), Mmus (\textit{Mus musculus}), Xtro (\textit{Xenopus tropicalis}), Cint (\textit{Ciona intestinalis}), Scer (\textit{Saccharomyces cerevisiae}).
\end{enumerate}

@@

%\begin{figure}[ht]
%  \includegraphics[scale=0.9]{figure-3/figure2.png}
%\end{figure}
%\newpage

>>>>

\section{Introduction}
Cilia are sensory organelles present on almost all cells in the human body
\citep{satir2010primary}.
They function primarily by providing sensory inputs to cells, but can
also function as motile propellers in certain cells
\citep{bloodgood2010sensory}.
Consistent with their importance in cell signalling,
defects in cilium formation are associated with genetic disorders
known collectively as ciliopathies, which can display
neurological defects \citep{valente2014primary},
skeletal abnormalities \citep{hammarsjo2017novel, waters2011ciliopathies},
and infertility \citep{inaba2016sperm} in addition to other phenotypes.

A large number of ciliopathies are linked to mutations in proteins that
localize to the cilium transition zone (referred to herein as TZ).
The TZ is the proximal-most region of the cilium and
acts as a diffusion barrier that regulates the
bi-directional transport of protein and lipid cargo at the base of the cilium
\citep{reiter2012base, szymanska2012transition}.
The TZ is essential for cilium formation (ciliogenesis) and cilium-associated
signalling.
The conserved TZ protein CEP290, for example, is mutated in at least
six different ciliopathies, and is essential for cilium formation in both humans
and Drosophila \citep{basiri2014migrating}.
Although the protein composition of TZs has been investigated in various
studies ranging from single gene analysis to proteomics,
processes underlying TZ maturation are not as well-studied.

In the Drosophila male germline, nascent TZs first assemble on basal bodies
during early G2 phase in spermatocytes and anchor directly to the plasma membrane.
This is followed by establishment of cilium-membrane connections
through a poorly understood mechanism
involving microtubule rearrangements \citep{gottardo2013cilium} and a
ciliary membrane that persists through meiosis \citep{riparbelli2012assembly}.
Ciliogenesis begins with the assembly of TZ proteins at the tips of
the nascent basal body.
This is followed by the anchoring of the TZ to the membrane and
establishment of a compartmentalized space bounded by the ciliary membrane
and the TZ.
TZ maturation has been described in
\textit{Paramecium} \citep{aubusson2015transition}
\textit{Caenorrhabditis elegans} \citep{serwas2017centrioles} and
\textit{Drosophila melanogaster} \citep{gottardo2013cilium}
and is most readily observed by an increase in TZ length from nascent to mature.
This are probably similar to the initial steps of TZ assembly and reorganization
in vertebrates.

Here, we show that the lipid phosphatidylinositol 4,5-bisphosphate (\PIP{})
is essential for proper TZ maturation in developing male germ cells in \textit{D. melanogaster}.
\PIP{} is a membrane lipid and one of seven phosphoinositides (PIPs)
present in eukaryotic cells.
It is primarily enriched at the plasma membrane and the base of cilia,
but its functions in ciliogenesis have not been studied.
We found that reduction of cellular \PIP{}
by expression of the bacterial PIP phosphatase SigD
or by inactivation of the PIP 5-kinase Skittles (Sktl) that localizes to cilia
induces longer than normal TZs without affecting TZ assembly.
These hyperelongated TZs show hallmarks of aberrant function similar to those
found in previously studied TZ mutants,
including loss of cilium-membrane associations.
We also find that the \textit{onion rings} allele of the
ciliopathy-associated \textit{EXOC8} gene encoding the human Exo84
decouples TZ hyperelongation from loss of cilium-membrane contacts.
Our results suggest that the plasma membrane lipid \PIP{} regulates TZ length,
maturation and function, acting to support ciliogenesis.

@@

\subsection{The \textit{onion rings} mutant decouples defects found in cells with reduced levels of \PIP{}}
A heterozygous mutation in the human \textit{EXOC8},
the human ortholog of yeast Exo84, is linked to
Joubert syndrome \citep{dixon2012exome}.
Exo84 is a component of the octameric exocyst complex, which binds
\PIP{} and regulates membrane trafficking at the plasma membrane.
We previously showed that male flies homozygous for the \textit{onion rings}
(\textit{onr}) mutant of Drosophila Exo84 were sterile, with phenotypes
similar to \sigd{} \citep{wei2008depletion}.
To investigate whether defects in TZ hyperelongation could be explained by
defective Exo84 function, we examined TZs in \textit{onr} mutants.
We found that the \textit{onr} mutant did not display hyperelongated TZs (Figure 5),
suggesting that Exo84 is dispensable for TZ maturation.

@@

\section{Discussion}
Cilia are essential for proper human development, and their disruption
can cause genetic disorders with severities ranging from congenital-
lethal \citep{}
to late onset \citep{}.
A large number of ciliopathies are associated with genes that encode for
proteins that function at the cilium transition zone (TZ).
Often, a single gene is linked to multiple ciliopathies with distinct phenotypes
and severities.
For example, \textit{CEP290} is linked to at least six different
cilium-associated disorders.
It is likely that these mutations do not completely abolish CEP290
function, and may be affected by genetic modifiers in an affected individual
\citep{zaghloul2010functional}.
Animal models with partial loss of TZ function may therefore provide
insight on intermediate phenotypes associated with human disease.
Here, we found that reduced levels of the membrane lipid \PIP{},
which is known to be enriched at the base of cilia,
affects TZ maturation, but not assembly,
leading to hyperelongated TZs that show signs of functional defects, and
may represent one such modifier for TZ function.

@@

%%
%    FIGURE 2
%
\textbf{Figure 2. Sktl is important for transition zone maturation and localizes to basal bodies.}
\begin{enumerate}[label={(\Alph*)}, nolistsep]
\item Expression of full-length Sktl suppresses \sigd{}-induced TZ hyperelongation.
\item Quantification of Cep290 and Ana1 lengths from control, \sigd{} and Sktl; \sigd{} ($n$\textgreater 60) from (A).
\item \textit{sktl}$^{2.3}$ clones exhibit TZ hyperelongation. Cilia are marked using
  Unc-GFP.
\item Phylogenetic tree of PIP5Ks showing conservation of cilium-associated functions.
  The scale bar represents expected amino acid substitutions per site and branch support values are shown in red (a value of 1 indicates maximum support).
Abbreviations: Cele (\textit{Caenorhabditis elegans}), Spur (\textit{Strongylocentrotus purpuratus}), Amel (\textit{Apis mellifera}), Aaeg (\textit{Aedes aegypti}), Dana (\textit{Drosophila ananassae}), Dmel (\textit{Drosophila melanogaster}), Hsap (\textit{Homo sapiens}), Mmus (\textit{Mus musculus}), Xtro (\textit{Xenopus tropicalis}), Cint (\textit{Ciona intestinalis}), Scer (\textit{Saccharomyces cerevisiae}).
\end{enumerate}

@@

%\begin{figure}[ht]
%  \includegraphics[scale=0.9]{figure-3/figure2.png}
%\end{figure}
\newpage
